coauthor_comments

### 2021-04-06 doc/manuscript/ForestGEO_treerings.Rmd
coauthor feedback

*(keep this paragraph mostly as is):*
Tree rings provide a long-term record of annual growth increments that is invaluable for understanding forests in an era of global change [@amoroso_dendroecology_2017; @fritts_dendroecology_1989]. 
Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterizing the interannual climate sensitivity of tree growth [@fritts_tree_1976; @braker_measuring_2002] and how it has changed through time [@sniderhan_growth_2016; @maxwell_declining_2016; @babst_twentieth_2019; @wilmking_global_2020; @au_demographic_2020].
Combined with forest censuses, they can be used to estimate forest woody productivity, $ANPP_{woody}$ [@graumlich_longterm_1989; @davis_forest_2009; @dye_comparing_2016; @teets_linking_2018] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019]. 
Tree rings also provide the long-term perspective critical to understanding how slowly changing environmental drivers including rising atmospheric carbon dioxide (CO~2~) concentrations, changing climate, and other anthropogenic and natural changes are influencing tree growth and forest productivity [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020].
This information is critical to predicting forest responses to anthropogenic changes, particularly climate change, and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@arora_carbon_2020].
Yet, collection and analysis of dendrochronological records has traditionally been optimized to detect climate signals rather than to predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbass-ahles_influence_2014; @klesse_sampling_2018;@babst_when_2018; @wilmking_global_2020].
As a result, prevailing approaches hold a number of limitations for using tree rings to address pressing questions concerning forest productivity in the current era of rapid environmental change.

@@

*(cut most of this paragraph? but JB really liked it):*
Dendrochronological methods to characterize the climate sensitivity of tree growth have been optimized to obtain the strongest possible climate signal for climate reconstructions, but face limitations when it comes to characterizing the long-term climate sensitivity of individual trees or $ANPP_{woody}$.
Traditional methods begin by fitting a function (commonly an exponential curve or a spline) to the growth record captured by each core, extracting residuals around the long-term trends, and then standardizing and averaging the residuals across cores to form a species-level ring-width index chronology [@fritts_tree_1976; @speer_fundamentals_2010].
Climate signals are then identified by examining month-by-month or seasonal correlations of the ring-width index chronology to one or more climate variables over the current year and, typically, previous growing season [e.g., @fritts_tree_1976; @meko_seascorr_2011; @zang_treeclim_2015].
Following identification of the top climate driver(s), statistical models describing their relationships to tree growth can be used for applications such as climate reconstruction [e.g., @buntgen_2500_2011] or projection of tree growth responses to climate change [e.g., @charney_observed_2016].
An important caveat for the latter, however, is that the slopes of correlations between climate variables and ring-width index chronologies are not identical to the mean slope of the relationship among individuals within the population, as the process of building species chronologies obfuscates individual-level responses [@carrer_individualistic_2011; @pederson_framework_2020].
Although population-level climate responses have been approximated based on climate sensitives derived from species chronologies [e.g., @charney_observed_2016; @helcoski_growing_2019], the removal of individual-level variation prior to analysis of climate sensitivity limits potential for using species chronologies to characterize the climate sensitivity of $ANPP_{woody}$.
In particular, analysis of species-level chronologies does not allow the direct characterization of known variation in climate sensitivity in relation to tree height, canopy position, or microhabitat characteristics such as topographic wetness index [e.g., @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_2021].

@@

*(keep mostly the same):*
Tree diameter at breast height (DBH) scales predictably with numerous traits affecting tree growth rate (e.g., height, crown size and position, root mass @enquist_global_2002; @niklas_plant_2004] and therefore is itself linked to growth [e.g., @foster_predicting_2016;  @muller-landau_testing_2006] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
However, for dendrochronological studies aimed at deciphering climate signals, DBH is not typically a variable of interest, and its influence, *along with the influence of stand dynamics in closed-canopy forests *, is removed through detrending [@cook_calculating_1997].
Moreover, many studies constrain sampling to only trees in canopy positions or those that appear to be old [@fritts_tree_1976; @huckaby_identification_2003; @pederson_external_2010].
While convenient for developing long chronologies and identifying climate signals [@fritts_tree_1976], this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity [@babst_when_2018].
Although climate correlations can be transformed to climate sensitivity [i.e., magnitude of response, *sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of ring width (RW) with DBH [@helcoski_growing_2019], they cannot be used to directly model known interactive effects of DBH and climate on tree growth [@trouillier_size_2019].
For example, drought sensitivity tends to increase with tree size [@bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020], and responses to temperature can also vary with tree size or canopy position [@rossi_agedependent_2007; @rollinson_climate_2021].
However, the removal of DBH signals via detrending makes it impossible to account for such size differences in climate sensitivity in a systematic, integrative way.
To use tree rings to predict tree growth, $ANPP_{woody}$, and forest dynamics, we need models that include DBH [e.g., @evans_fusing_2017; @klesse_amplifying_2020] and its potential climate interactions [e.g., @rollinson_climate_2021].

@@

For each species independently, we ran every combination of the candidate climate variables and DBH, including both first- and second-order terms of polynomial for each.
For climate response, we allowed concave-down fits, but ignored any concave-up fits on the basis that these are not expected biologically and, when identified in initial analyses, often appeared to be cases of over-fitting.

@@

To test for yeaar effects, we limited the analysis to species with reasonable coverage of the DBH x `year` matrix.
Specifically, we required that the species be represented by cores from  $\ge$ 3 trees spanning  $\ge$ 40% the total DBH range for $\ge$ 2/3 of the total time range analyzed. 
To avoid severe big-tree selection bias [@brienen_detecting_2012], we also required that the minimum DBH sampled be $\le$ 25 cm, if present.
Species that failed to meet these criteria (n= 8) were excluded from the analysis of temporal trends.
We then ran models as described above, including a  first-order linear effect of year.
We note that the random effect of tree should, in theory, avoid analytical biases arising from persistent growth differences among individuals [@brienen_detecting_2012; @brienen_tree_2017]. 
To verify that GLS model trends for yeaar were real, as opposed to an artifact of inherent covariation between DBH and yeaar within each core, we compared GLS results to an analysis of DBH-growth relationships by decade (Figs. `r decadal_analysis_figs_all`). 
For simplicity and consistency with previous literature, we present only the models with BAI as the growth metric, noting that responses were similar across growth metrics.

@@

*Growth rates changed directionally for the majority (`r percent_sig_neg_trend_of_analyzed`, depending on growth metric) of site-species combinations selected for analysis based on the sampling representation of DBH by yeaar (Fig. `r year_fig`).*
*These results were similar across growth metrics (Figs. `r all_GLS`), and consistent with an independent analysis of growth-DBH trends by decade (Figs. `r year_fig`, `r decadal_analysis_figs_all`), indicating that our approach is robust in its analytical structure.* (**EG: THESE SENTENCES (and the following) SOUND MORE LIKE RESULTS/ METHODS**)
Indeed, in a comparative analysis of several methods commonly used to detect growth trends, an approach **conceptually?** parallel (**AT comments that it doesn't seem parallel**) to that employed here (regional curve standardization) performed better at growth trend detection [@peters_detecting_2015] than two of the most commonly used methods for analyzing growth trends, conservative detrending and basal area correction [see @peters_detecting_2015], both of which would fail to capture the observed trends in RW and BAI with DBH (Fig. `r size`).
For instance, based on our finding that BAI tends to rise and fall with DBH, basal area correction would generally be biased towards finding positive trends for smaller trees and negative trends for larger trees.
However, our results remain subject to some potential sampling and survivorship biases that can influence overall trends in positive or negative directions [Fig. `r year_fig`, @brienen_detecting_2012; @groenendijk_no_2015; @brienen_tree_2017; @nehrbass-ahles_influence_2014].
While our analysis was designed to avoid some of the most severe potential biases [@brienen_detecting_2012], it is difficult, if not impossible, to control for all potential demography and survivorship biases, or to design sampling in a way that ensures unbiased representation of a species' growth rate at all points in the history of a stand [@brienen_detecting_2012; @brienen_tree_2017; @bowman_detecting_2013].
The observed trends should therefore be interpreted as representative of only the sampled trees, as opposed to all individuals of the species that existed throughout the time frame analyzed.
Within this context, signals of changing growth rate over time are attributable to some combination of stand dynamics (e.g., recruitment and succession, changing stand structure) and environmental drivers (e.g., climate drivers other than the primary drivers, indirect effects of climate change, rising atmospheric CO~2~, deposition of SO~2~ and NO~x~).

>>>>

*(keep this paragraph mostly as is):*
Tree rings provide a long-term record of annual growth increments that is invaluable for understanding forests in an era of global change [@amoroso_dendroecology_2017; @fritts_dendroecology_1989]. 
Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterizing the interannual climate sensitivity of tree growth [@fritts_tree_1976; @braker_measuring_2002] and how it has changed through time [@sniderhan_growth_2016; @maxwell_declining_2016; @babst_twentieth_2019; @wilmking_global_2020; @au_demographic_2020].
Combined with forest censuses, they can be used to estimate forest woody productivity [@graumlich_longterm_1989; @davis_forest_2009; @dye_comparing_2016; @teets_linking_2018] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019]. 
Tree rings also provide the long-term perspective critical to understanding how slowly changing environmental drivers including rising atmospheric carbon dioxide (CO~2~) concentrations, changing climate, and other anthropogenic and natural changes are influencing tree growth and forest productivity [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020].
This information is critical to predicting forest responses to anthropogenic changes, particularly climate change, and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@arora_carbon_2020].
Yet, collection and analysis of dendrochronological records has traditionally been optimized to detect climate signals rather than to predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbass-ahles_influence_2014; @klesse_sampling_2018;@babst_when_2018; @wilmking_global_2020].
As a result, prevailing approaches hold a number of limitations for using tree rings to address pressing questions concerning forest productivity in the current era of rapid environmental change.

@@

*(cut most of this paragraph? but JB really liked it):*
Dendrochronological methods to characterize the climate sensitivity of tree growth have been optimized to obtain the strongest possible climate signal for climate reconstructions, but face limitations when it comes to characterizing the long-term climate sensitivity of individual tree growth or stand productivity.
Traditional methods begin by fitting a function (commonly an exponential curve or a spline) to the growth record captured by each core, extracting residuals around the long-term trends, and then standardizing and averaging the residuals across cores to form a species-level ring-width index chronology [@fritts_tree_1976; @speer_fundamentals_2010].
Climate signals are then identified by examining month-by-month or seasonal correlations of the ring-width index chronology to one or more climate variables over the current year and, typically, previous growing season [e.g., @fritts_tree_1976; @meko_seascorr_2011; @zang_treeclim_2015].
Following identification of the top climate driver(s), statistical models describing their relationships to tree growth can be used for applications such as climate reconstruction [e.g., @buntgen_2500_2011] or projection of tree growth responses to climate change [e.g., @charney_observed_2016].
An important caveat for the latter, however, is that the slopes of correlations between climate variables and ring-width index chronologies are not identical to the mean slope of the relationship among individuals within the population, as the process of building species chronologies obfuscates individual-level responses [@carrer_individualistic_2011; @pederson_framework_2020].
Although population-level climate responses have been approximated based on climate sensitives derived from species chronologies [e.g., @charney_observed_2016; @helcoski_growing_2019], the removal of individual-level variation prior to analysis of climate sensitivity limits potential for using species chronologies to characterize the climate sensitivity of stand productivity.
In particular, analysis of species-level chronologies does not allow the direct characterization of known variation in climate sensitivity in relation to tree height, canopy position, or microhabitat characteristics such as topographic wetness index [e.g., @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_2021].

@@

*(keep mostly the same):*
Tree diameter at breast height (DBH) scales predictably with numerous traits affecting tree growth rate (e.g., height, crown size and position, root mass @enquist_global_2002; @niklas_plant_2004] and therefore is itself linked to growth [e.g., @foster_predicting_2016;  @muller-landau_testing_2006] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
However, for dendrochronological studies aimed at deciphering climate signals, DBH is not typically a variable of interest, and its influence, *along with the influence of stand dynamics in closed-canopy forests *, is removed through detrending [@cook_calculating_1997].
Moreover, many studies constrain sampling to only trees in canopy positions or those that appear to be old [@fritts_tree_1976; @huckaby_identification_2003; @pederson_external_2010].
While convenient for developing long chronologies and identifying climate signals [@fritts_tree_1976], this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity [@babst_when_2018].
Although climate correlations can be transformed to climate sensitivity [i.e., magnitude of response, *sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of stand productivity based on the scaling of ring width (RW) with DBH [@helcoski_growing_2019], they cannot be used to directly model known interactive effects of DBH and climate on tree growth [@trouillier_size_2019].
For example, drought sensitivity tends to increase with tree size [@bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020], and responses to temperature can also vary with tree size or canopy position [@rossi_agedependent_2007; @rollinson_climate_2021].
However, the removal of DBH signals via detrending makes it impossible to account for such size differences in climate sensitivity in a systematic, integrative way.
To use tree rings to predict tree growth, ecosystem productivity, and forest dynamics, we need models that include DBH [e.g., @evans_fusing_2017; @klesse_amplifying_2020] and its potential climate interactions [e.g., @rollinson_climate_2021].

@@

For each species independently, we ran every combination of the candidate climate variables and DBH, including both first- and second-order terms of polynomial for each.
For climate response, we allowed concave-down fits, but ignored any concave-up fits on the basis that these are not expected biologically.

@@

To test for yeaar effects, we limited the analysis to species with reasonable coverage of the DBH x `year` matrix.
Specifically, we required that the species be represented by cores from  $\ge$ 3 trees spanning  $\ge$ 40% the total DBH range for $\ge$ 2/3 of the total time range analyzed. 
To avoid severe big-tree selection bias [@brienen_detecting_2012], we also required that the minimum DBH sampled be $\le$ 25 cm, if present.
Species that failed to meet these criteria (n= 8) were excluded from the analysis of temporal trends.
We then ran models as described above, including a  first-order linear effect of year.
We note that the random effect of tree should, in theory, avoid analytical biases arising from persistent growth differences among individuals [@brienen_detecting_2012; @brienen_tree_2017]. 
To verify that GLS model trends for year were real, as opposed to an artifact of inherent covariation between DBH and year within each core, we compared GLS results to an analysis of DBH-growth relationships by decade (Figs. `r decadal_analysis_figs_all`). 
For simplicity and consistency with previous literature, we present only the models with BAI as the growth metric, noting that responses were similar across growth metrics.

@@

*Growth rates changed directionally for the majority (`r percent_sig_neg_trend_of_analyzed`, depending on growth metric) of site-species combinations selected for analysis based on the sampling representation of DBH by yeaar (Fig. `r year_fig`).*
*These results were similar across growth metrics (Figs. `r all_GLS`), and consistent with an independent analysis of growth-DBH trends by decade (Figs. `r year_fig`, `r decadal_analysis_figs_all`), indicating that our approach is robust in its analytical structure.* (**EG: THESE SENTENCES (and the following) SOUND MORE LIKE RESULTS/ METHODS**)
Indeed, in a comparative analysis of several methods commonly used to detect growth trends, an approach **conceptually?** parallel (**AT comments that it doesn't seem parallel**) to that employed here (regional curve standardization) performed better at growth trend detection [@peters_detecting_2015] than two of the most commonly used methods for analyzing growth trends, conservative detrending and basal area correction [see @peters_detecting_2015], both of which would fail to capture the observed trends in RW and BAI with DBH (Fig. `r size`).
For instance, based on our finding that BAI tends to rise and fall with DBH, basal area correction would generally be biased towards finding positive trends for smaller trees and negative trends for larger trees.
However, our results remain subject to some potential sampling and survivorship biases that can influence overall trends in positive or negative directions [Fig. `r year_fig`, @brienen_detecting_2012; @groenendijk_no_2015; @brienen_tree_2017; @nehrbass-ahles_influence_2014].
While our analysis was designed to avoid some of the most severe potential biases [@brienen_detecting_2012], it is difficult, if not impossible, to control for all potential demography and survivorship biases, or to design sampling in a way that ensures unbiased representation of a species' growth rate at all points in the history of a stand [@brienen_detecting_2012; @brienen_tree_2017; @bowman_detecting_2013].
The observed trends should therefore be interpreted as representative of only the sampled trees, as opposed to all individuals of the species that existed throughout the time frame analyzed.
For instance, ... *give an example/more clear explanation here? (see Camille's feedback)*
Within this context, signals of changing growth rate over time are attributable to some combination of stand dynamics (e.g., recruitment and succession, changing stand structure) and environmental drivers (e.g., climate drivers other than the primary drivers, indirect effects of climate change, rising atmospheric CO~2~, deposition of SO~2~ and NO~x~).

### 2021-04-06 doc/manuscript/ForestGEO_treerings_SI.Rmd
coauthor feedback

### Scotty Creek, Northwest Territories, Canada
The Scotty Creek site is a black spruce (*Picea mariana*) forest on rapidly thawing discontinuous permafrost [@sniderhan_growth_2016; @dearborn_permafrost_2020].
Vertical permafrost thaw has been causing drought stress in the shallow-rooted trees, resulting in growth declines [@sniderhan_growth_2016] and high mortality [@dearborn_permafrost_2020].

>>>>

### Scotty Creek, Northwest Territories, Canada
The Scotty Creek site is a black spruce (*Picea mariana*) forest on rapidly thawing discontinuous permafrost [@baltzer_forests_2014; @sniderhan_growth_2016; @dearborn_permafrost_2020].
Vertical permafrost thaw has been causing drought stress in the shallow-rooted trees, resulting in growth declines [@sniderhan_growth_2016] and high mortality [@dearborn_permafrost_2020].

### 2021-04-06 doc/manuscript/coauthor_feedback/2021_02/coauthor feedback.Rmd
coauthor feedback

## Jenn B
The SI description of the site looks good. Though you could pop in Baltzer et al. 2014 (https://onlinelibrary.wiley.com/doi/abs/10.1111/gcb.12349) in the citations for the first sentence of the Scotty appendix, if you are so inclined.

@@

- I'm not sure how useful it is to have ANPPwoody as an abbreviation of woody productivity: I've only seen it mentioned in the introduction (no equation etc) and it wouldn't take that much more space to have it spelled entirely. However, it's common enough that either way could work. 
- Table 1: I recommend the package "flextable" that is easy to use and super flexible (as expected from its name) when creating tables in Rmd, especially when compiling in word for which the package kableExtra doesn't work. 
- Table 2: perhaps indicate the share of plot biomass made up by the species included in the article to get an idea of their representativeness (although this is primarily a methodological article, representativeness of plot dynamics is not the main issue).
- L 602  I don't remember seeing individual growth rates in Meakem et al 2018 , but you know it better than me
-L314-316: if convex curves are likely overfitted, some concave ones could potentially be overfitted too?
- Discussion - Changing growth rates: perhaps I missed it, but one mechanism that could explain decreases in growth rates with time is the fact that the oldest trees, which are the only records available for the farthest years, are clear 'survivors/winners' and are thus likely to show higher than average growth rates: survival and growth are strongly linked (eg Aubry-Kientz et al 2015, https://doi.org/10.1002/ece3.1532). This is probably what you meant by 'survivorship biases' but I think this deserves an explanation

>>>>

## Jenn B (DONE)
The SI description of the site looks good. Though you could pop in Baltzer et al. 2014 (https://onlinelibrary.wiley.com/doi/abs/10.1111/gcb.12349) in the citations for the first sentence of the Scotty appendix, if you are so inclined.

@@

- I'm not sure how useful it is to have ANPPwoody as an abbreviation of woody productivity: I've only seen it mentioned in the introduction (no equation etc) and it wouldn't take that much more space to have it spelled entirely. However, it's common enough that either way could work.

**Removed abbreviation.**

- Table 1: I recommend the package "flextable" that is easy to use and super flexible (as expected from its name) when creating tables in Rmd, especially when compiling in word for which the package kableExtra doesn't work.

*https://ardata-fr.github.io/flextable-book/*

**Sounds great, and I want to use this alter, but for this paper I can't get it working quickly enough to be worth the time.**

- Table 2: perhaps indicate the share of plot biomass made up by the species included in the article to get an idea of their representativeness (although this is primarily a methodological article, representativeness of plot dynamics is not the main issue).

**I like the idea, but don't have that info readily available.**

- L 602  I don't remember seeing individual growth rates in Meakem et al 2018 , but you know it better than me

**in SI**

-L314-316: if convex curves are likely overfitted, some concave ones could potentially be overfitted too?

**removed the overfitting comment**

- Discussion - Changing growth rates: perhaps I missed it, but one mechanism that could explain decreases in growth rates with time is the fact that the oldest trees, which are the only records available for the farthest years, are clear 'survivors/winners' and are thus likely to show higher than average growth rates: survival and growth are strongly linked (eg Aubry-Kientz et al 2015, https://doi.org/10.1002/ece3.1532). This is probably what you meant by 'survivorship biases' but I think this deserves an explanation
